Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-27 Manuscript/manuscriptPS.tex
Add GROMOS-CKP POPC results with increasing amount of POPS, delivered by T. Piggot by email

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
    The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
    the error bar is given as the difference between the results divided by two. {\color{red} Why divided by two?}
  }
%  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
%    add GROMOS-CKP data into this plot.}
\end{figure}

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
%Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated" lipids might carry with them differing amounts of water between the different FFs.. -JJM

@@

Experimentally, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

>>>>

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
    the error bar is given as the difference between the results divided by two.
    % {\color{red} Why divided by two?}
    % Because then both values fit within the error bar.
  }
\end{figure}

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
% Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the
% specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated"
% lipids might carry with them differing amounts of water between the different FFs.. -JJM

@@

Experimentally, the both headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.
